---
title: "AdaptiveMCMCExample"
author: "Paul M"
date: "2/4/2021"
output: html_document
---

## Example adapted from manual of adaptMCMC

```{r, message=FALSE}
require(adaptMCMC)
require(coda)
set.seed(255)
NumberOfIterations <- 5000
```

We explore a 'Banana-shaped' distribution that has plenty of correlation:
```{r}
## log-pdf to sample from
p.log <- function(x) {
  B <- 0.03 # controls 'bananacity'
  -x[1]^2/200 - 1/2*(x[2]+B*x[1]^2-100*B)^2
}
```

Generate samples from that distribution using non-adaptive sampling.
We use the function MCMC, which is part of the adaptive MCMC package, and is designed to do adaptive MCMC.
```{r}
samp.1a <- MCMC(p.log, n=NumberOfIterations, init=c(0, 1), scale=c(1, 0.1),
               adapt=FALSE)
samp.1b <- MCMC(p.log, n=NumberOfIterations, init=c(0, 1), scale=c(1, 0.1),
               adapt=FALSE)
```

And now repeat, but this time using adaption:
```{r}
## 2) adaptive sampling
samp.2a <- MCMC(p.log, n=NumberOfIterations, init=c(0, 1), scale=c(1, 0.1),
               adapt=TRUE, acc.rate=0.234)
samp.2b <- MCMC(p.log, n=NumberOfIterations, init=c(0, 1), scale=c(1, 0.1),
               adapt=TRUE, acc.rate=0.234)
```

We plot the density that results and the samples we drew
```{r}
x1 <- seq(-15, 15, length=80)
x2 <- seq(-15, 15, length=80)
d.banana <- matrix(apply(expand.grid(x1, x2), 1, p.log), nrow=80)
par(mfrow=c(1,2))
image(x1, x2, exp(d.banana), col=cm.colors(60), asp=1, main="no adaption")
contour(x1, x2, exp(d.banana), add=TRUE, col=gray(0.6))
lines(samp.1a$samples[1:200,], type='b', pch=3)
image(x1, x2, exp(d.banana), col=cm.colors(60), asp=1, main="with adaption")
contour(x1, x2, exp(d.banana), add=TRUE, col=gray(0.6))
lines(samp.2a$samples[1:200,], type='b', pch=3)
```


And now we calcluate the Coda convergence diagnostics:
```{r}
## convert in object of class 'mcmc'
samp1a.coda <- convert.to.coda(samp.1a)
samp1b.coda <- convert.to.coda(samp.1b)
bothsamps1<-mcmc.list(samp1a.coda,samp1b.coda)
samp2a.coda <- convert.to.coda(samp.2a)
samp2b.coda <- convert.to.coda(samp.2b)
bothsamps2<-mcmc.list(samp2a.coda,samp2b.coda)

## ----------------------
## use functions of package 'coda'
gelman.plot(bothsamps1,main="No adaptation")
gelman.plot(bothsamps2,main="Adaptation")
print(gelman.diag(bothsamps1))
print(gelman.diag(bothsamps2))
```

We see an impressive improvement in performance when we use adaptive MCMC.
The price you pay for this usually is that the adaptive version runs more slowly.
For the problem here, this  price is minimal, but for complex datasets there can
be a substantial difference in run-time. Overall though, if you are having trouble
getting your MCMC analysis to converge, it is definitely worth considering trying
adaptive MCMC to see if it might help.

More exmaples of adaptativew MCMC can be found in the paper by Gareth Roberts that I 
have uploaded to the week 8 repository.
