---
title: "Tempering"
author: "Paul M"
date: "March 22, 2016"
output:
  pdf_document: default
  html_document: default
---

This example is based on one from http://neuromat.numec.prp.usp.br/sites/default/files/MCMC.pdf.
First we define a density to work with and plot it. We want something multi-modal
```{r}
library(mcmc)
mydens<-function(x) 0.5*dnorm(x)+0.2*dnorm(x,-8,2)+0.3*dnorm(x,15,3)
xx<-seq(-18,25,0.01)   # Some x-values to plot against
plot(xx,mydens(xx),type="l",xlab="x",ylab="Density",main="Our target density")
```

In this example we start from x=0 and use Metropolis-Hastings MCMC using a random walk proposal (standard deviation of 1, the default). We need to define a function to evaluate the log un-normalized posterior for metrop to work with.
```{r}
lupost<-function(x) log(0.5*dnorm(x)+0.2*dnorm(x,-8,2)+0.3*dnorm(x,15,3))
```
We next run a realization starting from 0 and using 1e5 steps: 
```{r}
set.seed(28651)
out<-metrop(lupost,initial=2,nbatch=1e5,scale=1)
```

We can see the evolution of the chain over time as follows:
```{r}
plot(out$batch[,1],1:1e5,type='l',xlab="x",ylab="idx",main="Simulated chain trajectory")
```
Let's construct a histogram to see what the posterior looks like
```{r}
hist(out$batch,breaks=500,probability=TRUE,xlab="x",ylab="Density",main="Estimated density compared to target")
lines(xx,mydens(xx),col=2,lwd=2)
```



We see that the chain had trouble mixing between the three peaks of the posterior, almost completely missing the righthand peak.
Now we try tempering, to see if we can improve on this.
We re-run the simulations using parallel tempering with 5 temperatures: 1, 0.7, 0.5, 0.3, 0.1. We can look at the target densities at each temperature with:

```{r}
witch.which <- c(0.1, 0.3, 0.5, 0.7, 1.0)   # these are the temperatures
ncomp <- length(witch.which)
my_cols <- c(4,1,1,1,2)
plot(xx,mydens(xx),type="l",lwd=2,col=2, xlab="x",ylab="Density",
     main="The five target densities")
sapply((ncomp-1):1,
       function(beta_idx) {
           beta <- witch.which[beta_idx]
           yy <- mydens(xx)^beta/integrate(function(x) mydens(x)^beta,-50,50,subdivisions = 1000)$value
           lines(xx,yy,lwd=2,col=my_cols[beta_idx])
       })
```

So, now let's run popuation MCMC in which we run 5 chains, one at each of those temperatures.
We use the "temper()" function in the MCMC R package. As part of this, we have to define which chains are 'neighbors' (and so can have their states swapped)

```{r}
neighbors <- matrix(FALSE, ncomp, ncomp)  # defines which chains are adjacent to which.
neighbors[row(neighbors) == col(neighbors) + 1] <- TRUE
neighbors[row(neighbors) == col(neighbors) - 1] <- TRUE
ludfun <- function(state) {
    stopifnot(is.numeric(state))
    stopifnot(length(state) == 2)
    # temper will pass this function arguments using the vector state (the current state)
    # the first elelment of the current state is the temperature of the calling chain
    icomp <- state[1]  
    stopifnot(icomp == as.integer(icomp))
    stopifnot(1 <= icomp && icomp <= ncomp)
    # now read off the current state of this chain (everything except the temperature)
    x <- state[-1]
    bnd <- witch.which[icomp]
    return(bnd*(log(0.5*dnorm(x)+0.2*dnorm(x,-8,2)+0.3*dnorm(x,15,3))))
}
thetas <- matrix(0, ncomp, 1)  # defines the initial thetas (=0 for each chain)
outREM <- temper(ludfun, initial = thetas, neighbors = neighbors, nbatch = 20000,blen = 1, scale = 1, parallel = TRUE, debug = TRUE)
```

Look at the results and compare it to what we get with 100000 steps (that is, the same computational cost) without tempering:

```{r}
hist(outREM$batch[,5,1],breaks=100,probability = TRUE,xlab="x",ylab="Density",main="Estimated densities compared to target")
out_hist <- hist(out$batch[1:100000,1],breaks=100,plot=FALSE)
lines(xx,mydens(xx),col=2,lwd=2)
lines(out_hist$mids,out_hist$density,col=4,lwd=2)
legend(10,0.15,c("Target","Est. Tempering","Est. No Tempering"),col=c(2,1,4),lwd=c(2,1,2),bty="n")
```

Let's do the same for the other temperature chains, to see what they look like:
```{r}
hist(outREM$batch[,4,1],breaks=100,probability = TRUE,xlab="x",ylab="Density",main="Estimated densities compared to target")
out_hist <- hist(out$batch[1:100000,1],breaks=100,plot=FALSE)
lines(xx,mydens(xx),col=2,lwd=2)
lines(out_hist$mids,out_hist$density,col=4,lwd=2)
legend(10,0.15,c("Target","Next warmest Temp","Est. No Tempering"),col=c(2,1,4),lwd=c(2,1,2),bty="n")

hist(outREM$batch[,3,1],breaks=100,probability = TRUE,xlab="x",ylab="Density",main="Estimated densities compared to target")
out_hist <- hist(out$batch[1:100000,1],breaks=100,plot=FALSE)
lines(xx,mydens(xx),col=2,lwd=2)
lines(out_hist$mids,out_hist$density,col=4,lwd=2)
legend(10,0.12,c("Target","Even warmer Temp.","Est. No Tempering"),col=c(2,1,4),lwd=c(2,1,2),bty="n")

hist(outREM$batch[,2,1],breaks=100,probability = TRUE,xlab="x",ylab="Density",main="Estimated densities compared to target")
out_hist <- hist(out$batch[1:100000,1],breaks=100,plot=FALSE)
lines(xx,mydens(xx),col=2,lwd=2)
lines(out_hist$mids,out_hist$density,col=4,lwd=2)
legend(10,0.07,c("Target","Very warm Temp.","Est. No Tempering"),col=c(2,1,4),lwd=c(2,1,2),bty="n")

hist(outREM$batch[,1,1],breaks=100,probability = TRUE,xlab="x",ylab="Density",main="Estimated densities compared to target")
out_hist <- hist(out$batch[1:100000,1],breaks=100,plot=FALSE)
lines(xx,mydens(xx),col=2,lwd=2)
lines(out_hist$mids,out_hist$density,col=4,lwd=2)
legend(10,0.053,c("Target","Warmest temp.","Est. No Tempering"),col=c(2,1,4),lwd=c(2,1,2),bty="n")

```

We can compare the trajectory of the chain's realization without tempering, with the one of the lowest temperature replica with tempering:

```{r}
layout(matrix(1:2,nc=2))
plot(out$batch[1:20000],1:20000,xlim=c(-18,25),xlab="x",ylab="idx",type="l",main="No tempering")
plot(outREM$batch[1:20000,5,1],1:20000,xlim=c(-18,25),xlab="x",ylab="idx",type="l",main="Tempering")
```

